---
title: "CUNY 607 HW7"
author: "Todd Weigel"
date: "October 16, 2016"
output: html_document
---

#Assignment Overview
###The assignment was to create an html, json, and xml "document" listing a series of books and attributes regarding them (name, title, etc.), and then be able to use R to parse through them and make dataframes from each file.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###The solution is straightforward, and I think the code below is self-explanatory (or as least so, with the code comments).  Regarding the question, as to whether the three dataframes are identical, I think the answer is "it depends".  Upon initial loading, no they were not, but they weren't wildly different, and the only real difference came through column names and rows names, which were essentially so the files match.  The only other difference was how the json file handled the authors data for the row that had multiple authors, but it wouldn't have taken much manipulation to make all entries the same.
###
```{r}
#bookChilds <- xmlChildren(root)

library(XML)
library(jsonlite)
library(stringr)


getDfFromXML <-function(filename)
{
    xml <- xmlParse(filename)
    root <- xmlRoot(xml)
    
    #get first "column" of data (eventually transposed to a row below)
    dfXML <- data.frame(xmlSApply(root[[1]], xmlValue))
    
    for (i in 2:xmlSize(root))
    {
        dfXML <- cbind(dfXML, data.frame(xmlSApply(root[[i]], xmlValue)))
    }
    #transpose, since rows were loaded as columns
    dfXML <- data.frame(t(dfXML),stringsAsFactors = FALSE)
    
    #just make rows numeric...rather than defaulted un-uesful names
    rownames(dfXML) <- seq(1:nrow(dfXML))
    
    return(dfXML)
}

getDfFromHTML <- function(filename)
{
    h <- htmlParse(filename)
    root <- xmlRoot(h)
    
    #pull header row which we will use as column names 
    header <- xpathSApply(root, "//table/tr/th", xmlValue)
    
    
    dfH <- data.frame(xpathSApply(root, "//table/tr/td[1]", xmlValue), stringsAsFactors = FALSE)
    #loop through all the table rows  (# note we make assumption the table we want will be first in page...)
    for(i in 2:length(xpathSApply(root, "//table/tr")))
    {
        dfH <- cbind(dfH, xpathSApply(root, str_c("//table/tr/td[", i, "]"), xmlValue))
    }
    colnames(dfH) <- header
    return(dfH)
}

getDfFromJSON <- function(filename)
{
    js <- fromJSON("books.json")
    dfJS <- data.frame(js, stringsAsFactors = FALSE)
    colnames(dfJS) <- str_replace(colnames(dfJS), "Uncoventional_Literature.", "") 
    return(dfJS)
}


dfBooksXML <- getDfFromXML("books.xml")
dfBooksHTML <- getDfFromHTML("books.html")
dfBooksJSON <- getDfFromJSON("books.json")

head(dfBooksXML)
head(dfBooksHTML)
head(dfBooksJSON)


```