---
title: "CUNY 607 HW9"
author: "Todd Weigel"
date: "October 16, 2016"
output: html_document
---

#Assignment Overview
####The assignment was to use the NY Times API to read in some data, and then transform that data in an R dataframe, that could be then queried as desired.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

####The basic functions to access the API and transform to a dataframe are below and are compact and simple.  And should hopefully be mostly self explanatory.  Comments will be given when we call these functions.

###
```{r}
library(RCurl)
library(stringr)
library(RJSONIO)
readApiKeyFromFile <-function(fileName)
{
    apiKey <- scan(fileName, what = "character")
}

readTimesAPI <- function(URL, key, q, pageNumber)
{
    params <- str_c(URL, "q=", q, "&page=", pageNumber, "&apiKey=", key)
    return(getURI(params))
}

parseJSON <-function (results)
{
    return(fromJSON(results, simplify = FALSE))
}

#This puts all the fields from a given row into a one row df
pullFields <- function(docRow) 
{           #rbind, gets around dataframe creation errors, even though we aren't joining anyrows together here
    as.data.frame(rbind(docRow[seq_along(1:length(docRow))]), stringsAsFactors = FALSE)
}

#This takes the passed in "document section" from the parsed JSON and creates a dataframe of it 
makeDF <- function(doc)
{
    
    df <- pullFields(doc[[1]])
    for(i in 2:length(doc))
    {
        df <- rbind(df, pullFields(doc[[i]]))
    }
    return(df)
}

```


####First We will populate some variables.  
* NY Times URL
* Read an my APIKEY from a file (so if rerunning this you will need to poulate a file with your personal API key.
* The query string we wish to perform a search for.
* The page number of query results we want to retrieve (the api returns 10 rows per "page"" of data)

```{r}
timesURL <- "https://api.nytimes.com/svc/search/v2/articlesearch.json?"
apiKey <- readApiKeyFromFile("timesAPIKey.txt")
query <- "brexit"
page <- 2
```


####Next we will call three functions to get a dataframe.  The first is to read the times api, using the variables we populated, and we will pass the results right into the parseJSON function which maniuplates the JSON from the API into an easier to work with format (a list).  Lastly call a function that (perhaps not efficiently) creates a dataframe by going row by row of the list and concatenating with RBind into a dataframe.

```{r}
parsed <- parseJSON(readTimesAPI(timesURL, apiKey, query, page))

df <- makeDF(parsed$response$docs)

```

####That is the entire program.  Lets run some commands to show that it actually works.  The below will show the head of the Dataframe, and then query a couple of the columns to show what was returned.

```{r}
head(df)

df$web_url[1:5]

df$news_desk[1]
```

####Lets finally rerun a different query for completeness, and show the data that is returned from that query.
```{r}

query <- "comey"
page <- 0
parsed <- parseJSON(readTimesAPI(timesURL, apiKey, query, page))

df <- makeDF(parsed$response$docs)
head(df)

df$web_url[1:5]

```

